Supplementary Information

### 2021-06-11 doc/manuscript/ForestGEO_treerings.Rmd
running head
final fixes

**Running Head:** [45 char]

@@

**Running headline:** [45 chars]

@@

Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), DBH, and calendar year on annual tree growth.
This approach allows us to address the above hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.

@@

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
---

@@

There was substantial variation across species-site combinations in the relationship between DBH and growth rate. (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, RW declined with DBH for a substantive portion of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`), contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production, @sillett_increasing_2010b; @sillett_comparative_2021] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that the approach presented here will prove valuable to understanding drivers of tree growth and forest change.
Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests wordlwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs. `r clim_sens`, `r interactions_fig`). 
Thus, the only way to analyze, understand, and predict global change effects is by simultaneously evaluating these effects. 
The method we present and apply here allows doing so.

@@

Code and full results are available via the project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: TBD) .
Data for [# TBD] of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, ... ; @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB; ...).

@@

`r decadal_analysis_verification_example_legend`

>>>>

@@

**Running head:** ecological analysis of global tree-ring data

@@

We address the above hypotheses (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude, using a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), DBH, and calendar year on annual tree growth.

@@

```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d; @dye_comparing_2016
---

@@

There was substantial variation across species-site combinations in the relationship between DBH and growth rate (Fig. `r size`).
Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
This variation was driven by two primary, interrelated factors: species ecology and stand history.
As hypothesized, RW declined with DBH for a substantive portion of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`). 
However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production, @sillett_increasing_2010b; @sillett_comparative_2021] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

@@

As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that the approach presented here will prove valuable to understanding drivers of tree growth and forest change.
Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests worldwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs. `r clim_sens`, `r interactions_fig`). 
Thus, the only way to analyze, understand, and predict global change effects on tree growth and forest productivity is by simultaneously evaluating these effects. 
The method we present and apply here allows doing so.

@@

Code and full results are available via the open-access project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: TBD).
Original data from each site are available through various public databases, as indicated in the original publications, or upon reasonable request from the authors.
Data for *[# TBD]* of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, *[others TBD]*, @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB).

@@

`r year_validation_legend`
